---
title: "WeatherInTheUSA"
output: html_document
---
# Severe Weather Events in the USA

## Synopsis

In this document the NOAA Storm Database is explored to understand the severe weather events in the USA and its economic consequences.

## Data Processing

Data for this document is taken from NOAA Storm Database available from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2
**readData()** function is used to read data from the website 

```{r}
library(lattice)

readData <- function(){
#====================================
#This function returns data required 
#==================================== 
  #download data from URL
  dataFileURL <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  #https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2
  #I am using Windows so https is removed and replaced with http curl does not work in my machine
  if (!file.exists(".\\data")){dir.create(".\\data")}#if there is no directory called data create one
  #now store downloaded file in data directory
  download.file(dataFileURL, destfile = ".\\data\\StormData.csv.bz2", mode ="wb")
  #mode set to "wb" because this is binary data
  #we do not need to use unzip as read.table and read.csv can directly read compressed files
 				
  ### READING data  
  stormData <- read.csv(".\\data\\StormData.csv.bz2", header = TRUE)
  stormData
}

stormData <- readData()

```
On my copy of the data downloaded on 22/06/2015, number of rows and columns are as follows:

Number of rows -
```{r}
nrow(stormData)
```
Number of columns -
```{r}
ncol(stormData)
```
There are various sever weather types identified by this dataset, to be precise there are 
```{r}
events <-unique(stormData$EVTYPE)
nrow(events)
```
unique sever weather types identified.

### Severe weather affecting population health
In exploring severe weather and its affect on population health I am first looking at event type and the number of fatalities.

```{r}
a <- aggregate(stormData$FATALITIES, by = list(stormData$EVTYPE), FUN = sum)
colnames (a) <- c("EVTYPE","TotalFatalities")
#sort by Fatalities (descending)
newdata <- a[order(-a$TotalFatalities),]
#remove non fatal events
n <-newdata[newdata$TotalFatalities > 0,]
colnames (n) <- c("EVTYPE","TotalFatalities")
```
The highest fatalities caused by sever weather are given below (only the events that caused more than 100 fatalities are presented)

```{r}
n[n$TotalFatalities >99,]
```
Below graph shows the top 5 sever weather events that causes high risk to life in the US

```{r}
top <- n[1:5,]
events <- top$EVTYPE
barplot(top$TotalFatalities, main = "Sever weather events and fatalities in the US", names.arg = events, col = "blue")
```
Below we look at sever weather and injuries
```{r}
ain <- aggregate(stormData$INJURIES, by = list(stormData$EVTYPE), FUN = sum)
colnames (ain) <- c("EVTYPE","totalInjuries")
#sort by INJURIES (descending)
newdatain <- ain[order(-ain$totalInjuries),]
#remove non injury events
nin <-newdatain[newdatain$totalInjuries > 0,]
colnames (nin) <- c("EVTYPE","totalInjuries")
```
The highest injuries caused by sever weather are given below (only the events that caused more than 100 injuries are presented)

```{r}
nin[nin$totalInjuries >99,]
```
Below graph shows the top 5 sever weather events that causes high risk to injuries/health in the US

```{r}
topin <- nin[1:5,]
events <- topin$EVTYPE
barplot(topin$totalInjuries, main = "Sever weather events and injuries in the US", names.arg = events, col = "red")

```



### Economic consequences of Severe Weather

In looking at economic consequences we observe crop damage and property damage caused by severe weather events in the USA

```{r}
subsetStorm <- stormData[c(8,25,27)]
names(subsetStorm) <- c("EVTYPE","PROPDMG","CROPDMG")
property<-aggregate(subsetStorm$PROPDMG,by = list (subsetStorm$EVTYPE), FUN = sum)
colnames(property) <- c("Event","PropertyDamage")
crop <- aggregate(subsetStorm$CROPDMG,by = list (subsetStorm$EVTYPE), FUN = sum)
colnames(crop) <- c("Event","CropDamage")

crop$PropertyDamage <- property$PropertyDamage
crop$TotalDamage <-rowSums(crop[,c("PropertyDamage","CropDamage")])
#order 
ordered <- crop[order(-crop$TotalDamage),]
total <- ordered[1:10,]

barplot(total$TotalDamage, main="Severe weather damage (Property and Crops)",
  xlab="Event", col=c("red"),  names.arg = total$Event)

```



## Resutlts

Above analysis showed that Tornados cause the highest fatalities and injuries in the USA. Therefore the importance of managing severe weather condition Tornados is highlighted.


